---
title: "Simulations In-Class Project"
date: "Due October 13, 2017 at 11:59pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:



With this project we will simulate a famous probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 


Since you all have completed problem 1, you first step will be to work through each of your groupmates code for problem #1 and comment on what is happening. Then as a team move forward in on direction as you move on to the rest of the problems 2-5

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet.
    
```{r}
#Nadia recommenting Kira's Code: 
#create a vector to contain the amounts of money the gabler has after each iteration of the game
#variables are bankroll(money left), num_bets(#of bets), and p(probability=0.5)
trials_to_bust <- function (bankroll, num_bets, p) {
  # create an empty vector to contain all the money left after each hand 
  #set the first element of this vector equal to the bankroll
  sum <- c(bankroll)
  
  # simulating a sequence of bets; adding a plus 1 since we begin step 1 with $1000
  for (bet in 2:(num_bets+1)) {
    if (sum[bet-1]!=0){ #if we have not gone "bust", continue the loop
      # +100 if rbinom(...) equals to 1, -100 if it equals to 0 (move up if we win and down if we lose)
      sum[bet] <- sum[bet-1] - 100*(-1)^rbinom(1, size=1, prob=p) 
    }else{
      # break the loop if the sum of money left is 0
      break
    }
  }
  
  # return the number of step where the gambler lost all money or 100 (if he still has money)
  return(c(bet=bet, bankroll=sum[bet-1]))
}

iterations <- 100000
hands <- 100
# vector containing results from all iterations of trials_to_bust (number of bet)
simulation <- replicate(iterations, trials_to_bust(1000,hands,0.5)["bet"])

# simulation[simulation<100] - all simulations where the gambler got busted before 100th bet
# proportion to the total number of iterations
Pr_get_busted <- length(simulation[simulation<=hands])/iterations

Pr_get_busted
```
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
```{r}
#Nadia recommenting parts of Kira's code

iterations <- 100000
hands <- 500 #only difference is changing the # of hands we play per itereation
# vector containing results from all iterations of trials_to_bust (number of bet)
simulation <- replicate(iterations, trials_to_bust(1000,hands,0.5)["bet"])

# simulation[simulation<100] - all simulations where the gambler got busted before 500th bet
# proportion to the total number of iterations
Pr_get_busted <- length(simulation[simulation<=hands])/iterations

Pr_get_busted
```
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
```{r}
#Nadia adding comments to Kira's code
iterations <- 100000
hands <- 5000 #new change in # hands
# vector containing results from all iterations of trials_to_bust (number of bet)
simulation <- replicate(iterations, trials_to_bust(1000,hands,0.5)["bet"])

# simulation[simulation<100] - all simulations where the gambler got busted before 100th bet
# proportion to the total number of iterations
mean_time_bust <- mean(simulation[simulation<=hands])

mean_time_bust
```
    d. the mean and variance of your bankroll after 100 hands (including busts).
```{r}
#Kira's code

iterations <- 100000
hands <- 100
# vector containing results from all iterations of trials_to_bust (bankrolls)
simulation <- replicate(iterations, trials_to_bust(1000,hands,0.5)["bankroll"])

# simulation[simulation<100] - all simulations where the gambler got busted before 100th bet
# proportion to the total number of iterations
mean(simulation)
var(simulation)
```
    e. the mean and variance of your bankroll after 500 hands (including busts).
```{r}
#Kira's code
iterations <- 100000
hands <- 500
# vector containing results from all iterations of trials_to_bust (bankrolls)
simulation <- replicate(iterations, trials_to_bust(1000,hands,0.5)["bankroll"])

# simulation[simulation<100] - all simulations where the gambler got busted before 100th bet
# proportion to the total number of iterations
mean(simulation)
var(simulation)
``` 
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

2. Repeat the previous problem with betting on black in American roulette, where the probability of winning on any spin is 18/38 for an even payout.
```{r}
#used Kira's code for Q2: all we need to do is to change p and changed 'hands' to 'spins'

trials_to_bust <- function (bankroll, num_bets, p) {
  # create a vector which will contain the amounts of money gambler has after each bet 
  # set the first element of this vector equal to the bankroll
  sum <- c(bankroll)
  
  # simulating a sequence of bets
  for (bet in 2:(num_bets+1)) {
    if (sum[bet-1]!=0){
      # +100 if rbinom(...) equals to 1, -100 if it equals to 0
      sum[bet] <- sum[bet-1] - 100*(-1)^rbinom(1, size=1, prob=p) 
    }else{
      # break the loop if the sum of money left is 0
      break
    }
  }
  
  # return the number of step where the gambler lost all money or 100 (if he still has money)
  return(c(bet=bet, bankroll=sum[bet-1]))
}

iterations <- 100000
spins <- 100
# vector containing results from all iterations of trials_to_bust (number of bet)
simulation <- replicate(iterations, trials_to_bust(1000,spins,18/38)["bet"])

# simulation[simulation<100] - all simulations where the gambler got busted before 100th bet
# proportion to the total number of iterations
Pr_get_busted <- length(simulation[simulation<=spins])/iterations

Pr_get_busted
```
3. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
```{r}
#commenting Frank's code 
game.Sim.Basic<- function(N){ #create a function for a simulation for the markov chain
Total=0 #begin with $0
  
game.hand <- rep(NA,N) #create an empty vector to store how much money you win at the end of each hand
Probability<-.48 #initial probability

for (i in 1:N){ #for each hand
  hand<-rbinom(1,1,Probability) #game simulation function to tell us if we either win or lose each game
  if(hand==1){#This is a win
    Total=Total+100 #gain $100 if you win
    Probability=Probability+.01 #add 1% to the probability if you win
      if(Probability>1){ #add a control to prevent your probability from going past 100%
        Probability=1
      }
  }else{
    Total=Total-100 #if you lose, deduct $100 from your total amount
    Probability=.48 #reset the probability
  }
}

return(Total) #return the total amount left at the end of the simulations

}

mean(replicate(100,game.Sim.Basic(100000)))
```
**Result is not a fair game as the amount left is not close to $0, which you would expect if you begin with $0 and if this is a fair game.** 
    b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
```{r}
game.Sim<- function(N,p){ #new variables are #of hands and starting probability
Total=0 
  
game.hand <- rep(NA,N) #create an empty vector to store how much money you win at the end of each hand
Probability<-p #initial probability (now undefined)

for (i in 1:N){ #for each hand
  hand<-rbinom(1,1,Probability) #game simulation function to tell us if we either win or lose each game
  if(hand==1){#This is a win; still move probability up by 1% if a win
    Total=Total+100
    Probability=Probability+.01
      if(Probability>1){
        Probability=1
      }
  }else{
    Total=Total-100 #still lose $100 and reset probability to original probability (still undefined)
    Probability=p
  }
}

return(Total) #still returning the total amount of $$$ left over after 

}


p <-.49 #start at .49 because we know that .48 is unfair and .50 is unfair, so starting at .49 is the mid of that 
pmax<-.50 #set max and min so that our p does not go beyond those bounds
pmin<-.46
FinalPrevious<-  1000000000000000000000 #Arbitrialiary large Starting Value so we don't bust

while(1){ #tells r to keep running while loop
Final<-mean(replicate(100,game.Sim(100000,p)))#Simulation reduced to allow program to run - will start at 100000 though

#print(paste("Current p is ",p,"while amount lost/gained is ",Final ))  #statement to see simulation at work

if(Final<0) #if the $ leftover if less than 0
{
  pmin<-p
  p<-p+(pmax-p)/2 #Increment by 1/2 of the difference if too large
}else if (Final>0){
  pmax<-p
  p<-p-(p-pmin)/2 #Increment by 1/2 of the difference if too small
}




if(abs(FinalPrevious)<abs(Final) & abs(FinalPrevious)<100){ #This looks for the tipping point in our chain where our increment made things worse instead of better, we also want to give some portion of what we consider "small enough" here we use 1
  print(paste("Final P is ",p," while the amount loss/gained is ",FinalPrevious))
  break
}

FinalPrevious <- Final

}

```
    c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 
```{r}
#same variables but we will change how we move our p up or down until we reach a fair p
game.Sim<- function(N,p,increment){ 
Total=0
  
game.hand <- rep(NA,N)
Probability<-p

for (i in 1:N){
  hand<-rbinom(1,1,Probability)
  if(hand==1){#This is a win
    Total=Total+100
    Probability=Probability+increment #will be defined later how to increase our p by a certain increment
      if(Probability>1){
        Probability=1
      }
  }else{
    Total=Total-100
    Probability=p
  }
}

return(Total)

}
    

p <-.48
increment<-.01
incrementMin<-0 #prvents our p value from going below 0
incrementMax<-1 #prevents our p value from going above 1

FinalPrevious<-  1000000000000000000000 #Arbitrialiary large Starting Value

while(1){
Final<-mean(replicate(100,game.Sim(10000,p,increment)))#Simulation # reduced to allow program to run

#print(paste("Current increment is ",increment,"while amount lost/gained is ",Final )) #statement to see simulation at work

if(Final<0) #same but instead of changing p, we are changing the increment we are moving by
{
  incrementMin<-increment
  increment<-increment+(incrementMax-increment)/2 #Increment by 1/2 of the difference if too large
}else if (Final>0){
  incrementMax<-increment
  increment<-increment-(increment-incrementMin)/2 #Increment by 1/2 of the difference if too small
}




if(abs(FinalPrevious)<abs(Final) & abs(FinalPrevious)<1000){ #This looks for the tipping point in our chain where our increment made things worse instead of better, we also want to give some portion of what we consider "small enough" here we use 1000.  This is changed from before because the increment affects our valus less then p did
  print(paste("Final Increment is ",increment," while the amount loss/gained is ",FinalPrevious))
  break
}

FinalPrevious <- Final

}


```

4. Creating a Bootstrap function. There is a particular concept called [bootstrapping]
(https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) where we can easily create 95% confidence intervals, even for complex estimators.

The steps of this process are:

  a. Draw a sample, with replacement, from your data which is the same length of your data.
  b. Calculate the statistic of interest on this boostrap sample (ie mean, variance, regression,...)
  c. Peform steps 1:2 at least 1000 times over until you have a vector of your statistics. 
  d. The lower bound of a 95% CI will be the 0.025 percentile
  e. The upper bound of a 95% CI will be the 0.975 percentile
  
  Make a function called `boot_ci` which calculates the 95% confidence interval in this manner. 
  
**We did the above all at once with the below function**
```{r}
boot_ci <- function(df) {
  #create an empty df to store stata (means, vars, etc...)
  stata_df <- data.frame()
 
  for (i in 1:1000){ #repeating 1000 times
   #sample rows with replacement #4a
    my_sample <- df[sample(x = nrow(df), size = nrow(df), replace=TRUE), ] 
   
   #applying "mean" to the 2nd dim (columns) #4b
    stata <- apply(my_sample,2,mean) 
    
   #add a new stata row to stata_df
    stata_df <- rbind(stata_df, stata)
  }

  #now we need to find CI for each column of stata_df given percentiles
  #create a df of CIs
  CI_df <- data.frame()
  
  for (j in 1:ncol(stata_df)) {
    #calculate margin of error based in bounds
    m <- mean(quantile(stata_df[,j], c(0.975,0.025))) 
    
    #calculate mean given bounds and margin error
    mar <- quantile(stata_df[,j], 0.975) - m
    
    CI_df <- rbind(CI_df, list(m, "+/-", mar))
  }
  
  names(CI_df) <- c("mean", "+/-", "margin of error")
  rownames(CI_df) <- colnames(df)
  return(CI_df)
}

my_data <- replicate(100, game.Sim.Basic(10000))
boot_ci(my_data)
```



5. For problems 3b and 3c, you calculated a mean value. Because you saved these final results in a vector, use the bootstrap to estimate the variance of the return in each case for your final answer. Once you have these results, which game has the smaller variance in returns?