---
title: "Simulations In-Class Project"
date: "Due October 13, 2017 at 11:59pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:



With this project we will simulate a famous probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 


Since you all have completed problem 1, you first step will be to work through each of your groupmates code for problem #1 and comment on what is happening. Then as a team move forward in on direction as you move on to the rest of the problems 2-5

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
    ```{r}
#create function to only store a 0 or 1 if we busted or still have money remaining after a number of bets 
gamblersRuin <- function(nn, upprob=0.5, start.here=1000) {
  output <- rep(NA, nn)
  output[1] <- start.here
  for(iteration in 2:nn+1) {
    output[iteration] <- output[iteration-1] + 200*rbinom(1,1,upprob) - 100 #iteration-100 to move up or down by 100
    
    if(output[iteration]==0){
      return(0)
      break} 
    
  }
  return(output[length(output)]) #return value of money left after playing after a number of bets
  
}

probBust <- function(nn, N){
bust <- rep(NA, N) #create an empty vector to run 1000 simulations
for (i in 1:N){
  output = gamblersRuin(nn) #run randomwalk for only 100 hands 
  if (output >0) { 
  bust[i] = 0     #save value of 0 if we did have money after playiing 100 hands
  } else {
    bust[i] = 1 #save value of 1 if we did not have money after playing 100 hands
  }
}
return(mean(bust)) #mean of all busts (1) over all the 100 hands
}

probBust(100, 5000) #probability for playing 100 hands in 5000 iterations of gamblersRuin
```
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
    ```{r}
probBust(500, 5000)
```
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
    ```{r}
gamblersRuin.bust <- function(nn, upprob=0.5, start.here=1000) {
  output <- rep(NA, nn)
  output[1] <- start.here
  for(iteration in 2:nn) {
    output[iteration] <- output[iteration-1] + 200*rbinom(1,1,upprob) - 100 #iteration-100 to move up or down by 100
    
    if(output[iteration]==0){ #add break and store as 0 when you bust
      return(output[1:iteration]) #tells us where in the iteration we went bust
      break}
  }
}

mean.bust <- function(n = 1000){ #refers to 1000 iterations
  output <- rep(NA, n)
  for(i in 1:n){
    play <- gamblersRuin.bust(5000) #create vector of play to store each iteration of the game until we bust (for 5000 hands played)
    output[i] <- length(play) #calculate the length of play to get the position we went bust
  }
  return(mean(output)) #calculate the mean of the output to get the average time we went bust
}
mean.bust()
```
    d. the mean and variance of your bankroll after 100 hands (including busts).
    ```{r}

#keep original gamblersRuin to maintain return value of money
gamblersRuin.last <- function(nn=100, upprob=0.5, start.here=1000) { #nn is the number of hands you play per iteration
  output <- rep(NA, nn)
  output[1] <- start.here
  for(iteration in 2:(nn+1)) {
    output[iteration] <- output[iteration-1] + 200*rbinom(1,1,upprob) - 100 #move up or down by 100
    
    if(output[iteration]==0){
      return(0)
      break} 
    
  }
  return(output[length(output)]) #return value of money left 
  
}

#create function to calculate mean and variance of bankroll
mean.var <- function(n = 5000) { #refers to 1000 iterations
  output <- rep(NA, n)
  for(i in 1:n){
    play <- gamblersRuin.last(100) #play gambler's ruin 100 times
    output[i] <- play
  }
  return(c(mean(output), var(output)))
}
mean.var()
```
    e. the mean and variance of your bankroll after 500 hands (including busts).
    ```{r}
mean.var <- function(n = 5000) { #refers to 1000 iterations
  output <- rep(NA, n)
  for(i in 1:n){
    play <- gamblersRuin.last(500) #play gambler's ruin 100 times
    output[i] <- play
  }
  return(c(mean(output), var(output)))
}
mean.var()

```
 
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

2. Repeat the previous problem with betting on black in American roulette, where the probability of winning on any spin is 18/38 for an even payout.
```{r}
trials_to_bust <- function (bankroll, num_bets, p) {
  # create a vector which will contain the amounts of money gambler has after each bet 
  # set the first element of this vector equal to the bankroll
  sum <- c(bankroll)
  
  # simulating a sequence of bets
  for (bet in 2:(num_bets+1)) {
    if (sum[bet-1]!=0){
      # +100 if rbinom(...) equals to 1, -100 if it equals to 0
      sum[bet] <- sum[bet-1] - 100*(-1)^rbinom(1, size=1, prob=p) 
    }else{
      # break the loop if the sum of money left is 0
      break
    }
  }
  
  # return the number of step where the gambler lost all money or 100 (if he still has money)
  return(c(bet=bet, bankroll=sum[bet-1]))
}

iterations <- 5000
hands <- 100
# vector containing results from all iterations of trials_to_bust (number of bet)
simulation <- replicate(iterations, trials_to_bust(1000,spins,18/38)["bet"])

# simulation[simulation<100] - all simulations where the gambler got busted before 100th bet
# proportion to the total number of iterations
Pr_get_busted <- length(simulation[simulation<=spins])/iterations

Pr_get_busted
```


3. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
    ```{r, eval=FALSE}
#function constants are playing 100,000 hands and repeat simulation 99 times
#we will still go up and down by $100
#want to play one hand at a time, observe if the output increased by $100. If it did, then increase the probability by 1%. If we lost $100, then reset probability to 48%
# markov.chain <- function(x, p = 0.48){ #x is the number of hands we play
#   outcomes <- rep(NA, x)
#   outcomes <- 0
#   for(i in 2:x) {
#     output[i] <- output[i-1] + 200*rbinom(1, 1, p) - 100
#   
#   if(diff(ouput[[i]] > 100) {
#     return(probStart + 0.01)
#   } else {
#     return(probStart == 0.48)
#   }

#commenting Frank's code 
game.Sim<- function(N){ #create a function for a simulation for the markov chain
Total=0 #begin with $0
  
game.hand <- rep(NA,N) #create an empty vector to store how much money you win at the end of each hand
Probability<-.48 #initial probability

for (i in 1:N){ #for each hand
  hand<-rbinom(1,1,Probability) #game simulation function to tell us if we either win or lose each game
  if(hand==1){#This is a win
    Total=Total+100 #gain $100 if you win
    Probability=Probability+.01 #add 1% to the probability if you win
      if(Probability>1){ #add a control to prevent your probability from going past 100%
        Probability=1
      }
  }else{
    Total=Total-100 #if you lose, deduct $100 from your total amount
    Probability=.48 #reset the probability
  }
}

return(Total) #return the total amount left at the end of the simulations

}

mean(replicate(100,game.Sim(100000)))
```
    b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
```{r}
game.Sim<- function(N,p){ 
Total=0 
  
game.hand <- rep(NA,N) 
Probability<-p

for (i in 1:N){
  hand<-rbinom(1,1,Probability)
  if(hand==1){#This is a win
    Total=Total+100
    Probability=Probability+.01
      if(Probability>1){
        Probability=1
      }
  }else{
    Total=Total-100
    Probability=p
  }
}

return(Total)

}

#We start p at .485 since we know that .48 is unfair and .50 is unfair (but in the boon of the player) 
p <-.49 #start at .49 because we know that .48 is unfair and .50 is unfair, so starting at .49 is the mid of that 
pmax<-.50 
pmin<-.46
FinalPrevious<-  1000000000000000000000 #Arbitrialiary large Starting Value

while(1){ #tells r to keep running while loop
Final<-mean(replicate(100,game.Sim(10000,p)))#Simulation # reduced to allow program to run

print(paste("Current p is ",p,"while amount lost/gained is ",Final ))

if(Final<0) #if the $ leftover if less than 0
{
  pmin<-p
  p<-p+(pmax-p)/2 #Increment by 1/2 of the difference if too large
}else if (Final>0){
  pmax<-p
  p<-p-(p-pmin)/2 #Increment by 1/2 of the difference if too small
}




if(abs(FinalPrevious)<abs(Final) & abs(FinalPrevious)<100){ #This looks for the tipping point in our chain where our increment made things worse instead of better, we also want to give some portion of what we consider "small enough" here we use 1
  print(paste("Final P is ",p," while the amount loss/gained is ",FinalPrevious))
  break
}

FinalPrevious <- Final

}

```
    
    c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 
```{r}

```
  


4. Creating a Bootstrap function. There is a particular concept called [bootstrapping]
(https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) where we can easily create 95% confidence intervals, even for complex estimators.

The steps of this process are:

  a. Draw a sample, with replacement, from your data which is the same length of your data.
```{r}

```

  b. Calculate the statistic of interest on this boostrap sample (ie mean, variance, regression,...)
  c. Peform steps 1:2 at least 1000 times over until you have a vector of your statistics. 
  d. The lower bound of a 95% CI will be the 0.025 percentile
  e. The upper bound of a 95% CI will be the 0.975 percentile

Make a function called `boot_ci` which calculates the 95% confidence interval in this manner. 

5. For problems 3b and 3c, you calculated a mean value. Because you saved these final results in a vector, use the bootstrap to estimate the variance of the return in each case for your final answer. Once you have these results, which game has the smaller variance in returns?